Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-09-26 paper1_jiturra.Rmd
update

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.5'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention to the link between class relations and redistributive preferences in contemporary societies. However, economic inequality unevenly affects class relations and their influence on support for redistribution across societies. While those in the upper and middle classes reinforce their privileged access to resources and diverse networks, the working class has been disproportionately affected by social exclusion as inequality rises. Consequently, inequality consolidates segregated lifeworlds as it increases the divide between the lives and experiences of the different social classes. Meeting people from diverse class positions can help individuals become more aware of other people's lifestyles and worldviews. Homogeneous upper-class networks may reduce empathy and solidarity towards those in need, resulting in decreased support for redistribution. Conversely, segregated lower-class social networks can increase marginalization and reinforce redistribution support. This study explores the relationship between class-based network segregation and redistributive preferences, employing cross-national data from 31,694 individuals in 31 societies. The main findings suggest that social class conditions the association of network homogeneity on redistributive preferences, where working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity decreases support for redistribution. In addition, the conditional influence of network segregation weakens in unequal societies, especially for the upper classes. Implications for the study of class relations and political attitudes are discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

 During the past decades, cross-national studies on redistributive preferences and social class have predominantly focused on individuals and households [@lindh_class_2020]. In contrast, less attention has been paid to the role of social networks on redistributive preferences. Sociologists have demonstrated that individuals tend to form segregated social networks according to social class [@mcpherson_birds_2001;@otero_lives_2022]. Additionally, attitudes toward economic inequality can be affected by class segregation, as it implies distance from other social classes'  economic conditions, lifestyles, and worldviews [@lindh_missing_2021;@lee_consider_2023;@paskov_crossclass_2022]. Therefore, focusing on social ties can improve the understanding of the relationship between social class and redistributive preferences, as it broadens the scrutiny of class relations in terms of internalization of shared norms, class identity formation, and collective solidarity. [@Svallfors2006;@goldthorpe_service_1982]. In this discussion, the first part of this study examines the micro-level relationship between class-based network segregation and redistributive preferences. Subsequently, I focus on how cross-national differences in income inequality can moderate the relation of network segregation and support for redistribution.

While previous research has predominantly examined the impact of social class through an individualist lens, more attention has recently been paid to the role of social ties in class relations. This omission is especially surprising since social classes are rooted in social relations of production, which makes them inherently relational in their economic foundations and the power dynamics involved in distributive struggles [@wright_comparative_1989]. Moreover, the normative basis of class relations introduces the relevance of dimensions of solidarity and reciprocity, which are argued to provide the moral foundation for legitimacy and popular support for welfare schemes [@mau_moral_2003]. Thus, beyond the influence of individual class positions, recent efforts have highlighted that, while the class positions of network ties affect redistributive preferences, segregation in homogeneous class-based social settings can make attitudes even more pronounced [@otero_lives_2022].

As network segregation can widen the class divide in redistributive preferences, little is known about how income inequality may affect the association between class-based network segregation and support for redistribution. On the one hand, studies on the class-attitude link suggest that income inequality is crucial for understanding how class relations are reflected in redistributive demands, as it represents the current state of distributive class struggle in contemporary capitalist societies [@edlund_democratic_2015;@curtis_how_2015]. A consistent finding in this literature is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, while the already high redistributive demands of the working class remain stable regardless of inequality levels [@sachweh_why_2019;@dimick_altruistic_2017]. On the other hand, income inequality has implications for social participation and network composition. Indeed, inequality reinforces stratified access to social activities and widens the distance between classes, resulting in a higher diversity of networks and activities in the upper classes and more marginalization of the lower classes [@lancee_income_2012;@otero_differences_2023;@pichler_patterns_2007]. However, current efforts to examine the impact of income inequality have focused primarily on social networks or support for redistribution. Thus, a key question that remains unanswered is the extent to which higher income inequality moderates the relationship between social class-based network segregation and redistributive preferences.

To the best of my knowledge, no research has comparatively studied the role of economic inequality in the relationship between class-based network segregation and redistributive preferences. Therefore, this study aims to answer two key questions:

(1) How does class-based network segregation affect redistributive preferences?
(2) To what extent does economic inequality moderate the relationship between class-based network segregation and redistributive preferences?

In this study, I used a sample of 31,694 individuals from 31 countries from the International Social Survey Program (ISSP) 2017. This dataset provides unprecedented cross-national comparative data on social networks, social class, and attitudes toward redistribution.

@@

Over the past decades, the study of political attitudes in industrialized societies has consistently demonstrated the relevance of social class as a driver of public opinion [@lindh_class_2020, p. 421]. In this view, social class not only represents individuals' labor market relations but also their economic interests and moral perspectives on the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Here, redistributive preferences are understood as the support for policies and mechanisms aimed at reducing economic inequality through the redistribution of wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier to less wealthy individuals or groups in society [@garcia-sanchez_two_2022].

Class-based explanations of redistributive preferences have primarily focused on the individual or household level but are not exclusively limited to these contexts. It has been argued that redistributive preferences can be explained through labor market situations, which encompass access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009]. Additionally, while material interests may dominate in situations of scarcity, value-driven motivations become a significant explanatory factor for redistributive preferences under conditions of greater certainty and are weaker under material precariousness [@Kulin2013;@Maldonadoetal2019]. Other approaches suggest that the substantial time workers spend on their jobs allows social relations in the workplace to imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. For example, the continuous and diverse social interactions inherent in interpersonal services can foster empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, vertical monitoring in managerial occupations and the emphasis on autonomy in self-employed roles tend to strengthen self-interested and conservative political views [@oesch_electoral_2018].

Empirically, the current understanding of class divides in redistributive preferences is extensive [@lindh_public_2015;@curtis_how_2015;@langsaether_more_2020;@brooks_why_2010]. Nevertheless, considering class-based network ties highlights the relevance of the class situation, as it encompasses economic interests and life chances that drive communal action beyond individuals [@weber_class_2011, pp. 57-59]. Hence, considering the structure of social ties allows one to look more comprehensively into how class relations structure redistributive preferences.

@@

Class relations can be understood as the structure of social ties between different classes within the social structure, represented by networks formed by various social classes [@blau_macrosociological_1977]. Empirical evidence consistently shows homophily in social relations within the social network literature [@mcpherson_birds_2001]. This indicates that friendships and family ties are often homogeneous regarding social status and demographic characteristics, whereas distant ties connect individuals to different social groups and enhance network diversity [@diprete_segregation_2011;@lazarsfeld_friendship_1954]. Moreover, socialization preferences do play a role in the formation of segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, the similarity of attitudes within these networks arises more from the social differentiation structuring class ties than from individual preferences [@feld_focused_1981].

The study of class relations from a network perspective can be approached in two ways. First, network diversity is defined as the degree of connectedness to dissimilar occupations, representing vertical access to resource diversity within social networks [@lin_access_1986]. In this context, diversity refers to the variability of different ties within the network without necessarily providing a reference point to describe network similarity. Second, network segregation, defined as the absence of cross-class connections, indicates how similar network ties are in reference to the individual. This perspective is conceptually closer to homophily, as it is anchored in individual class positions and has been empirically addressed through network homogeneity [@otero_open_2021].

How does social class stratify social networks? Empirical studies on network diversity reveal that higher civic engagement in formal organizations increases the likelihood of the upper class forming connections with diverse individuals, contrasting with the more homogeneous participation observed among the working class [@pichler_social_2009]. Similar patterns are found in the composition of social ties, where the upper and intermediate classes maintain more diverse and prestigious social environments compared to the working class [@cepic_how_2020; @carrascosa_class_2023]. In contrast, studies on network segregation suggest that property-based boundaries are less permeable than authority-based ones. For example, @wright_relative_1992 suggest that class interests widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their higher contact frequency with manual workers make the formation of friendship ties more likely. Similarly, @otero_open_2021 found that the intermediate class exhibits greater permeability compared to the more homogeneous networks of the working class. This suggests that limited life chances and lower capacity for social engagement ultimately result in segregation. Conversely, the upper class is more homogeneous because it tends to self-select, a practice that ultimately seeks to reproduce its privileged positions [@otero_open_2021].

## Network segregation and attitudes toward redistribution

Besides individualistic class-based mechanisms in redistributive preferences, the argument that network ties have implications for attitude formation is not novel. In particular, two approaches have discussed the role of social relations on redistributive preferences: reference groups and class-based networks.

@@

The second approach suggests that social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' social class but also by the class positions of their network ties [@paskov_crossclass_2022]. Thus, social influence processes can either align or divide opinions depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These arguments reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences [@morris_attenuation_1996]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, segregated interactions may lead individuals to perceive the lives of different classes as more distant and undermine feelings of social integration [@sachweh_moral_2012]. Thus, segregation can potentially diminish social cohesion as it exacerbates perceptions of others as strangers which in turn reduces empathy and solidarity [@otero_lives_2022].

@@

In summary, as homogeneity refers to the overall degree of segregation without distinguishing between class positions I anticipate a weak direct association between network homogeneity and redistributive preferences. Conversely, I hypothesize that the association of network homogeneity with redistributive preferences is conditional on social class as homogeneous social networks should reinforce attitude similarity (*segregation hypothesis*). Specifically, I propose that greater network segregation in the lower classes is associated with higher redistributive preferences, whereas greater segregation in the upper classes is related to lower redistributive preferences. Therefore, the first hypothesis is as follows:

> H1: _The greater the network segregation in the lower (upper) classes, the higher (lower) the redistributive preferences. _

## Economic inequality as context for class relations and redistributive preferences

Studies on economic inequality, social class, and redistributive preferences indicate that social classes react differently to rising economic inequality. Theoretically, political economists have suggested that high-income individuals are far from monolithic in their redistributive preferences, arguing that their concerns about the harmful _consequences_ of economic inequality (e.g., crime) ultimately motivate altruistic support for redistribution [@dimick_models_2018;@dimick_altruistic_2017;@rueda_externalities_2016]. Conversely, the moral economy literature in sociology has argued that the differences among the affluent can be explained as a matter of distributive justice evaluations about the _procedures_ for resource allocation [@liebig_sociology_2016].

Empirically, it has been shown that affluent groups are more responsive as they perceive the overall opportunity structure and social mobility chances are affected by rising inequality [@sachweh_why_2019]. Likewise, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of justice in the conditions for getting ahead [@kim_socioeconomic_2018]. In contrast, low-income individuals perceive ascribed characteristics as more important in constraining the opportunity structure, regardless of current income inequality [@sachweh_why_2019]. In a similar perspective, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class has a poor political meaning, making distributive conflicts weakly institutionalized around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can also be explained as they are often more aware of the incentives and societal consequences of income inequality [@curtis_how_2015;@Svallfors2006].

How does income inequality affect social networks? As income inequality implies stronger barriers to establishing trustworthy social connections, this also has consequences on the structure of class relations. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] suggest that experiencing marginalization is deeply associated with diminished chances for participating openly in social life, which can be exacerbated in contexts of greater inequality. Also, societies with higher income inequality can erode social relations as they exhibit reduced social trust that operates as the basis for mutual understanding and feelings of solidarity [@kragten_income_2017]. Likewise, in contexts of lower income inequality and encompassing welfare states, higher levels of civic and social participation increase generalized trust and reduce social conflict [@uslaner_inequality_2005]. In a similar vein, @yamamura_social_2012 showed that redistributive preferences rise as community-level participation in social activities increases. Moreover, high-income citizens in these communities are specifically more likely to support collective solidarity through government redistribution.

Empirically, cross-national studies show that already-stratified access to social activities and diverse networks is further reinforced in unequal societies. For instance, @letki_getting_2015 show that extensive networks in unequal societies are more common among low-income individuals who rely more on close ties for social support than high-income individuals. Additionally, @pichler_social_2009 demonstrated that class differences in participation in civic and family networks are exacerbated in more unequal societies. Likewise, @lancee_income_2012 found that income-based stratification in civic participation becomes more pronounced as inequality increases. Also, @otero_differences_2023 found that economic inequality enhances stratified access to network diversity, suggesting that it amplifies the interdependence among cultural, economic, and social capital. Thus, the upper classes can navigate diverse social settings while preserving their distinctive position, whereas the lower classes often remain marginalized and segregated due to the choices of others [@otero_open_2021, p.24].

To summarize, I expect that income inequality weakens the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class (*segregation hypothesis*) will be less pronounced in contexts of higher income inequality (*mitigation hypothesis*). Given the above considerations, the second hypothesis reads as follows:

> H2: _The greater the income inequality, the weaker the conditional association of network segregation by social class on redistributive preferences._

<!-- A simplified framework of the hypotheses is shown in Figure X. -->

@@

I use two indicators to measure redistributive preferences. The first indicator corresponds to support for government redistribution and it is measured by the item: "It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes." The second indicator is egalitarian preferences and it is measured by the item: "For a society to be fair, differences in people's standard of living should be small." Both indicators use a five-point Likert scale with the following categories: 'Strongly agree' (1), 'Agree' (2), 'Neither agree nor disagree'(3), 'Disagree' (4), and 'Strongly disagree'(5). Following @svallfors_government_2013, I reverse-coded, averaged and normalized the indicators in a 0 to 100 index, where higher values reflect stronger redistributive preferences.

@@

To measure social class, I employ the Erikson-Goldthorpe-Portocarrero (EGP) class scheme [@Erikson1992]. The EGP scheme is one of the most consistent and validated measures for social class positions in comparative research and has demonstrated its validity in both industrialized and late-industrialized societies [@evans_political_2013;@barozet_measurement_2021]. Therefore, information about occupations, self-employment status, and the number of employees is used to classify respondents into six class positions. Following previous research, a simplified version of the EGP class scheme that collapses three classes is employed [@sosnaud_class_2013;@edlund_influence_2003]. Specifically, this version distinguishes among the Service Class (higher and lower managerial and professionals), Intermediate Class (routine nonmanual workers and self-employed), and Working Class (manual supervisors, skilled and unskilled manual workers) (see Table \@ref(tab:tab-an-classag) in the Appendix).

I employed the position generator as the basis for the class-based network homogeneity measure. This instrument has been widely used in social capital studies and follows an ego-centered approach where is assumed that social ties to different hierarchical positions in the social structure provide access to social resources [@vandergaag_position_2008]. Here, a list of ten occupations is displayed and declared as a "Family or relative," "Close friend", "Someone else I know," or "No one." The first three categories are binary coded as "Knows" (1) and the rest as "Does not know" (0). Then, the sum of all occupations known represents the total number of ties per respondent.

>>>>

---
title: "Class-based network segregation, Economic Inequality and Redistributive Preferences across societies"
css: "input/css/custom.css" # custom CSS para html
fontsize: 12pt
linestretch: '1.5'          # interlineado  
#linestretch: '2.0'          # interlineado 
papersize: a4
link-citations: yes         # citas linkeadas
lang: en-US
# date: "`r format(Sys.time(), '%A %d %B %Y %X')`"
date: "`r format(Sys.Date(), '%d %b %Y')`"
author:
- name: Julio Iturra-Sanhueza
  affiliation: Bremen International Graduate School of Social Sciences
  email: jiturra@uni-bremen.de
abstract: | 
  | Rising economic inequality has increased attention on the link between class-based networks and redistributive preferences in contemporary societies. Previous research has shown that class differences in redistributive preferences tend to narrow in more unequal societies, primarily due to higher redistributive preferences within the service class. However, little is known about how economic inequality affects the relationship between class-based networks and redistributive preferences. Moreover, income inequality can reinforce the role of social class in shaping social networks, highlighting how the lower class becomes increasingly segregated from other classes. Experiencing homogeneous class-based networks leads to segregated lifeworlds, which widen the gap between social classes and reduce awareness of the living conditions and worldviews of others. Empirically, I expect that homogeneous upper-class networks may reduce empathy and solidarity toward those in need, resulting in lower support for redistribution. Conversely, homogeneous working-class networks increase marginalization and strengthen the demand for redistribution. At the country level, I anticipate that in more unequal societies, the relationship between class-based network segregation and redistributive preferences weakens, due to the narrower differences between the upper and lower classes. This study examines the relationship between class-based network segregation and redistributive preferences and how it is moderated by economic inequality, utilizing cross-national data from 31,694 individuals across 31 societies. In line with these expectations, the findings suggest that social class shapes the association between network homogeneity and redistributive preferences: working-class homogeneity drives stronger redistributive preferences, while greater upper-class homogeneity reduces support for redistribution. Additionally, the conditional relationship between network segregation and social class weakens as inequality increases, particularly in the service class. The implications for studying class-based networks and political attitudes are also discussed.
  | 
  | **Keywords**: social networks, segregation, social class, income inequality, redistributive preferences 
output:
  bookdown::pdf_document2:
    template: null
    toc: false
    keep_tex: false
    number_sections: true
    pandoc_args:
      - --template=input/mytemplate.tex #custom template para usar autores con afiliacion  
  bookdown::html_document2:
    toc: false
    code_folding: hide
    number_sections: false
    toc_float:
      collapsed: false
      smooth_scroll: false

@@

Over the past few decades, cross-national studies on social class and redistributive preferences have predominantly centered on individual class positions [@lindh_class_2020]. Recently, however, a growing body of research has begun to focus on how the class profile of social networks influences redistributive preferences [@lindh_missing_2021; @lee_consider_2023; @paskov_crossclass_2022]. Although prior studies have shown that class-based network segregation—defined as the concentration of similar-class network ties—can undermine attachment to society [@otero_lives_2022], there is still limited understanding of how being embedded in a homogeneously class-based network impacts attitudes toward inequality and redistribution [@otero_power_2023]. Contributing to this debate, this research first aims to investigate the individual-level relationship between class-based network segregation and redistributive preferences. In addition, it has been suggested that social classes tend to form segregated networks according to the homophily principle, understood as the higher chances of forming social ties with others with similar characteristics [@mcpherson_birds_2001;@otero_lives_2022]. Furthermore, cross-national comparisons indicate that class-based network stratification strengthens in societies with high economic inequality, which in turn drives stronger social distance between classes, particularly among the lower classes [@otero_differences_2023; @pichler_social_2009]. Against this backdrop, the present study seeks to contribute to understanding how cross-national differences in income inequality act as a moderating factor in the relationship between class-based network segregation and redistributive preferences.

While segregation may deepen existing class divisions in redistributive preferences [@paskov_crossclass_2022], little is known about how income inequality shapes this relationship. On the one hand, research on the class-attitude link suggests that income inequality is crucial for understanding how class relations—i.e., the social and political distance between classes—translate into redistributive demands, as it reflects the state of class struggle in contemporary capitalist societies [@edlund_democratic_2015; @curtis_how_2015]. A consistent finding is that the upper classes tend to be more supportive of redistribution in contexts of high inequality, whereas the already high redistributive demands of the lower classes remain relatively stable, regardless of inequality levels [@sachweh_why_2019; @dimick_altruistic_2017]. On the other hand, it is well-documented that income inequality affects social relations, reinforcing stratified access to social activities and widening the distance between classes. In such conditions, the upper classes tend to be more socially active and maintain more diverse networks, while the lower classes become increasingly inactive and segregated as inequality rises [@lancee_income_2012; @otero_differences_2023; @pichler_social_2009]. However, current research efforts have primarily examined the impact of income inequality moderates the relations of social class on either social networks or redistributive preferences. Thus, this study aims to address two key questions:

(1) How does class-based network segregation influence redistributive preferences?
(2) To what extent does economic inequality moderate the relationship between class-based network segregation and redistributive preferences?

In this research, I use a sample of 31,694 individuals from 31 countries, drawn from the 2017 International Social Survey Program (ISSP). This dataset offers unprecedented cross-national comparative data on social networks, social class, and attitudes toward redistribution.

@@

Over the past few decades, research on political attitudes in industrialized societies has consistently highlighted the significance of social class as a key driver of public opinion [@lindh_class_2020]. Social class, in this context, not only reflects individuals' labor market relations but also their economic interests and moral perspectives regarding the role of the market and the state in the distribution and redistribution of resources [@Svallfors2006]. Redistributive preferences refer to the support for policies and mechanisms aimed at reducing economic inequality by redistributing wealth and resources [@mccall_americans_2009]. These preferences encompass views on taxation, welfare programs, public services, and other government interventions designed to transfer resources from wealthier individuals or groups to those with fewer resources [@garcia-sanchez_two_2022].

Empirically, the class divide in redistributive preferences is well documented [@lindh_public_2015; @curtis_how_2015; @langsaether_more_2020; @brooks_why_2010]. Class-based explanations of redistributive preferences have predominantly focused on individuals. According to self-interest-driven theories, economic resources or exposure to risk explain why working classes with fewer resources and greater job insecurity tend to support redistribution more than the upper classes [@meltzer_rational_1981; @rehm_risks_2009]. Furthermore, while material interests often dominate in conditions of scarcity, value-based motivations, such as egalitarianism, may drive stronger support for redistribution under conditions of greater security and weaken under material hardship [@Kulin2013; @Maldonadoetal2019].

Other approaches emphasize the role of social relations in the workplace, which can imprint normative views that ultimately shape political opinions due to the significant time people spend at work [@oesch_redrawing_2006]. For instance, continuous and diverse social interactions in interpersonal service roles can foster empathy and reinforce egalitarian values [@kitschelt_occupations_2014]. Conversely, vertical oversight in managerial positions and the emphasis on autonomy in self-employed roles often bolster self-interested and conservative political views [@oesch_electoral_2018].

Given this context, studying the connections between social classes from a network perspective offers a comprehensive way to understand redistributive preferences beyond the individual level. By focusing on network ties, the collective dimension of class relations is brought to the forefront, emphasizing that social ties encompass both social and economic resources [@lin_access_1986], which serve as the foundation for communal action [@weber_class_2011, pp. 57-59].

@@

Theoretically, class relations can be understood as the structure of social ties between different classes within the broader social system, represented by networks spanning various social strata [@blau_macrosociological_1977]. Research in social network literature consistently demonstrates that homophily—the tendency for individuals to associate with others who are similar—is a structured and persistent feature of social relations [@mcpherson_birds_2001]. For instance, friendships and family ties often display homogeneity in terms of social status or demographic characteristics, while more distant ties tend to connect individuals to different social groups, thereby contributing to network diversity [@diprete_segregation_2011; @lazarsfeld_friendship_1954]. Moreover, it is well established that socialization preferences play a role in forming segregated networks [@visser_attitudes_2004; @homans_human_1951]. However, this research aligns with the prevailing view that attitudinal similarity within networks arises primarily from structural contact opportunities shaped by the class composition of social ties, rather than from homophilic preferences for socializing with like-minded individuals [@feld_focused_1981].

To investigate the implications of networks for redistributive preferences, I propose distinguishing between two closely related—but distinct—perspectives on how class-based networks manifest within social structures.

First, the concept of _network diversity_ refers to the degree of connectedness to dissimilar ties (e.g., different occupations or activities), representing vertical access to diverse resources within social networks [@lin_access_1986]. In this context, diversity reflects the dispersion of different ties (alteri) across the network, independent of the individual’s (ego) characteristics [@otero_power_2023]. Empirical studies, such as those by @pichler_social_2009 and @lancee_income_2012, argue that higher civic engagement in formal organizations increases the likelihood of the upper classes forming connections with more diverse individuals, whereas the working classes tend to engage in more homogeneous participation. Similar patterns emerge in social tie composition, where the upper and intermediate classes maintain more diverse and prestigious social environments than the working class [@cepic_how_2020; @carrascosa_class_2023].

The second and primary perspective adopted in this study is _network segregation,_ defined as having contact primarily with people of the same class. This approach measures the similarity between an individual’s (ego) network ties and is conceptually closer to homophily, as it is rooted in an individual’s class position and has been empirically examined through network homogeneity [@otero_open_2021]. Evidence suggests that property-based boundaries are far less permeable than authority-based ones in the formation of cross-class ties. For example, @wright_relative_1992 suggests that class interests tend to widen the social distance between proprietors and manual workers, while the intermediate position of supervisors and their more frequent contact with manual workers make cross-class friendships more likely. Similarly, @otero_open_2021 identified a U-shaped pattern of acquaintance network segregation in Chile, where the middle classes (e.g., lower managerial professionals, clerks, and manual supervisors) exhibit lower network homogeneity. Likewise, segregation is most prevalent on both ends of the class structure, particularly among the lower classes. They argue that segregation in the lower classes is often driven by limited life chances and reduced social participation, while upper-class segregation is largely explained by self-selection—a practice that reinforces privileged positions, which can coexist with a broader range of social connections [@otero_open_2021].

Given this context, one theoretical implication is that experiencing segregated class-based networks may be related to class differences in redistributive preferences. However, whether such segregation—empirically understood as network homogeneity—intensifies or mitigates class differences in these preferences remains a question to be empirically addressed.

## Network segregation and redistributive preferences

In addition to rational self-interest or value-based explanations, the argument that social ties have implications for attitude formation is not precisely new. More specifically, two approaches have discussed the role of social relations in redistributive preferences: reference groups and class-based networks.

@@

This research adopts the second approach, which suggests that social networks provide a comprehensive picture of the class relations that contribute to group identity formation and internalization of social norms [@kalmijn_social_2007]. Specifically, it has been argued that redistributive preferences are influenced not only by individuals' social class but also by the class positions of their network ties [@paskov_crossclass_2022]. Thus, opinions can either align or divide through social influence processes depending on the class positions of contacts and the level of network segregation [@lindh_missing_2021]. These arguments reflect the notion that classes are characterized as collectivities with varying degrees of cohesion and solidarity, comprising asymmetric status-based interactions related to material resources, cultural practices, and political preferences [@morris_attenuation_1996]. Following @sachweh_moral_2012, social integration can be impeded in societies with few opportunities for contact between different social classes, creating an "empathy gulf" that hinders individuals from understanding others' lifestyles amid rising inequality. Consequently, segregated interactions may lead individuals to perceive the lives of different classes as more distant and undermine feelings of social integration [@sachweh_moral_2012]. Thus, segregation can potentially diminish social cohesion as it exacerbates perceptions of others as strangers which in turn reduces empathy and solidarity [@otero_lives_2022].

@@

In summary, I anticipate a weak direct association between network segregation and redistributive preferences. This is mainly because network homogeneity - as the proportion of similar class network ties - does not distinguish between anchor class positions and refers to the overall degree of segregation. Conversely, I hypothesize that the association of network homogeneity with redistributive preferences is conditional on social class as homogeneous social networks should reinforce attitude similarity (*segregation hypothesis*). Specifically, I propose that greater network segregation in the lower classes is associated with higher redistributive preferences, whereas greater segregation in the upper classes is related to lower redistributive preferences. Therefore, the first hypothesis is as follows:

> H1: _The greater the network segregation in the lower (upper) classes, the higher (lower) the redistributive preferences._

## Economic inequality as context for class-based segregation and redistributive preferences

The literature on the role of economic inequality on (1) the class-attitude link and (2) social relations have been independently developed.

Theoretically, studies on economic inequality, social class, and redistributive preferences show varying class reactions to rising inequality. Political economists suggest that high-income individuals are far from being monolithic in their redistributive preferences, motivated by concerns about inequality's adverse effects (e.g., crime), leading to altruistic support for redistribution [@dimick_models_2018; @dimick_altruistic_2017; @rueda_externalities_2016]. Conversely, the moral economy literature in sociology explains affluent differences through evaluations of distributive justice regarding resource allocation procedures [@liebig_sociology_2016]. At the same time, the literature on the role of income inequality on social cohesion has argued that unequal societies strengthen barriers to trustworthy social connections. Neckerman and Torche [-@neckerman_inequality_2007, p. 344] noted that marginalization is linked to reduced participation in social life, exacerbated by greater inequality. Also, rising economic inequality erodes social relations by reducing social trust, which underpins mutual understanding and collective solidarity [@kragten_income_2017]. Conversely, it is noted that societies with low inequality and comprehensive welfare states enhance civic and social participation, increasing generalized trust and reducing social conflict [@uslaner_inequality_2005]. Similarly, it has been argued that social participation can nurture higher chances for cross-class interactions and motivate more egalitarian attitudes, being upper-class citizens who support stronger collective solidarity through government redistribution in more cohesive communities [@yamamura_social_2012].

Empirically, it has been shown that affluent groups are more responsive, perceiving that rising inequality affects the overall opportunity structure and social mobility chances [@sachweh_why_2019]. Also, higher perceived inequality of opportunity among the upper classes can motivate support for redistribution as a matter of procedural justice [@kim_socioeconomic_2018]. In contrast, low-income individuals view ascribed characteristics as more crucial in constraining opportunities, irrespective of current income inequality [@sachweh_why_2019]. Similarly, @edlund_democratic_2015 argue that in unequal societies with residual welfare states, social class lacks significant political meaning, weakening distributive conflicts around traditional working-class organizations (e.g., unions). Thus, stronger support for state-organized redistribution among the upper classes can be attributed to their greater awareness of the incentives and societal consequences of income inequality [@curtis_how_2015; @Svallfors2006]. At the same time, cross-national comparative evidence indicates that stratified access to social activities and networks is reinforced in unequal societies. For example, @pichler_social_2009 demonstrated that class differences in civic and family network participation are exacerbated in unequal societies. Similarly, @lancee_income_2012 found that income-based stratification in civic participation increases with inequality. Also, @otero_differences_2023 found that economic inequality enhances 
interdependence among cultural, economic, and social capital, which results in stronger stratified chances for cross-class interactions. Thus, in unequal societies, the upper classes can navigate diverse social settings while maintaining their distinct positions, whereas the lower classes remain marginalized and segregated due to others' choices [@otero_open_2021, p.24].

Against this backdrop, there are at least three emerging theoretical implications on the role of income inequality on class-based networks and redistributive preferences. The first implication is that it is plausible to assume that income inequality drives greater segregation, as the result of lower participation and social trust. However,  it is worth noticing that resource interdependency seems to exacerbate social exclusion for the lower classes, while the upper classes may hold more cohesive experiences that drive stable or even lower segregation in more unequal societies. A second implication is that the relationship between class and attitudes weakens in highly unequal societies, where the upper classes tend to show more support for redistribution, contrasting with the relatively stable preferences of the lower classes. In addition, it is also true that in egalitarian societies with more encompassing welfare states class-based distributive struggles tend to be more institutionalized (e.g. through unions or political parties). As a result, class positions are more closely implied in political conflict as the institutional framework allows them to better articulate their interests. A third implication is that as inequality increases and the class divide in redistributive preferences narrows, this may also be extended to the influence of class-based segregation on redistributive preferences. As the class-attitude link becomes mild in more unequal contexts, it is also expected that homogeneous class-based networks might not have the same strength where the role of class positions loses significance. Therefore, I expect income inequality to weaken the interaction between network homogeneity and social class. Specifically, I hypothesize that the conditional relationship between network homogeneity and social class stated at the individual level (*segregation hypothesis*) will be less pronounced in the context of higher income inequality (*mitigation hypothesis*). Given these considerations, the second hypothesis is as follows:

> H2: _The greater the income inequality, the weaker the conditional association of network segregation by social class with redistributive preferences._

A simplified framework of the hypotheses is shown in Figure \@ref(fig:scheme).

```{r scheme, echo=FALSE, out.width='80%',fig.cap='Hypotheses at country and respondent levels'}
knitr::include_graphics('output/images/paper1-scheme.png')
```

@@

I use two indicators to measure redistributive preferences. The first indicator corresponds to support for government redistribution and it is measured by the item: "It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes." The second indicator is egalitarian preferences and it is measured by the item: "For a society to be fair, differences in people's standard of living should be small." Both indicators use a five-point Likert scale with the following categories: 'Strongly agree' (1), 'Agree' (2), 'Neither agree nor disagree'(3), 'Disagree' (4), and 'Strongly disagree'(5) (*r* = 0.62). Following @svallfors_government_2013, I reverse-coded, averaged and normalized the indicators in a 0 to 100 index, where higher values reflect stronger redistributive preferences.

@@

To measure social class, I employ the Erikson-Goldthorpe-Portocarrero (EGP) class scheme [@Erikson1992]. The EGP scheme is one of the most consistent and validated measures for social class positions in comparative research and has demonstrated its validity in both industrialized and late-industrialized societies [@evans_political_2013;@barozet_measurement_2021]. Therefore, information about occupations, self-employment status, and the number of employees is used to classify respondents into six class positions. Following previous research, a simplified version of the EGP class scheme that collapses three classes is employed [@sosnaud_class_2013;@edlund_influence_2003]. Specifically, this version distinguishes among the Service Class (higher and lower managerial and professionals), Intermediate Class (routine nonmanual workers and self-employed), and Working Class (manual supervisors, skilled and unskilled manual workers) [^egpclass] (see Table \@ref(tab:tab-an-classag) in the Appendix).

[^egpclass]: Self-employed farmers (IVc) are grouped in the self-employed class, while agricultural labor (VIIb) is in the working class.

I employed the position generator as the basis for the class-based network homogeneity measure. This instrument has been widely used in social capital studies and follows an ego-centered approach where is assumed that social ties to different hierarchical positions in the social structure provide access to social resources [@vandergaag_position_2008]. The position generator employed included a list of ten occupations. Here, occupations are presented to the respondent in a grid that allows them to declare whether they know (or not) a person who performs that occupation. The tie can be clasified in four groups, defining tie to this person is as a "Family or relative," "Close friend,” "Someone else I know," or "No one.”  With this information, the first three categories are coded as 1 to represent the presence of a tie ("Knows") and 0 as the abscence of a conection to a person with that occupation ("Does not know").  Subsequently, all declared ties were summed to represent the total number of occupations known by the respondent.
